Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-08-20 jso.qmd
add the discussion section

The index properties, including smoothness and squintability, offer
numerical metrics to characterise the complexity of projection pursuit
optimisation problems. @tbl-smoothness-squintability presents the
parameters for calculating both metrics estimated from the Gaussian
process (variance, range, smooth, and nugget) and the scaled logistic
function ($\theta_1$ to $\theta_4$) for each case considered in the
simulation. The column "smooth" is used as the smoothness metric and the
column "squint" is calculated as equation (@eq-squintability-parametric)
as the squintability metric. @tbl-mod-output presents the results of
fitting a generalised linear model with a binomial family and a logit
link function to a sample of data in @tbl-mod-data, where all the three
components (success rate, JS hyper-parameters, and index properties) are
combined. The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the index
properties: smoothness and squintability. Specifically, using 10 more
jellyfish and 10 more tries increases the odd ratio of success by
`r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`% and
`r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%,
respectively. However, being flagged with long runtime and an increase
of data dimension reduce the success rate by
`r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and
`r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension` are significant, suggesting
their importance relative to JSO hyper-parameters in the optimisation
success. In defining a projection pursuit problem, factors such as the
shape-to-find, the index function used, and the data dimension,
determine the properties such as smoothness and squintability. These
metrics can then be compared across different problems to understand
their relative complexites. The results suggest that squintability has a
more significant influence than smoothness on the success rate of JSO.
Once the characteristics of the projection pursuit problem are fully
understood, increasing JSO hyper-parameters can enhance the search
effectiveness, however, it is also important to consider the resulting
increase in computational complexity.

@@

>>>>

The index properties, including smoothness and squintability, offer
numerical metrics to characterise the complexity of projection pursuit
optimisation problems. @tbl-smoothness-squintability presents the
parameters for calculating both metrics estimated from the Gaussian
process (variance, range, smooth, and nugget) and the scaled logistic
function ($\theta_1$ to $\theta_4$) for each case considered in the
simulation. The column "smooth" is used as the smoothness metric and the
column "squint" is calculated as equation (@eq-squintability-parametric)
as the squintability metric. @tbl-mod-output presents the results of
fitting a generalised linear model with a binomial family and a logit
link function to a sample of data in @tbl-mod-data, where all the three
components (success rate, JS hyper-parameters, and index properties) are
combined. The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the index
properties: smoothness and squintability. Specifically, using 10 more
jellyfish and 10 more tries increases the odd ratio of success by
`r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`% and
`r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%,
respectively. However, being flagged with long runtime and an increase
of data dimension reduce the success rate by
`r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and
`r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension` are significant, suggesting
their importance relative to JSO hyper-parameters in the optimisation
success.

@@

# Discussion {#sec-discussion}

The two properties, smoothness and squintability, can be calculated once factors such as the target shape, the chosen index function, and the data dimension are defined in a projection pursuit problem. In practice, these metrics can be used to characterise the complexity of the optimisation task and inform the optimiser selection. For example, a large smoothness value suggests a smooth index function and both gradient-based and random sampling optimisers tend to perform well. A large squintability value indicates a distinct difference in index values between optimal regions and others. When one jellyfish enters the optimal region, the high index value it generates will be clearly distinguishable from spurious values due to this pronounced difference and lead other jellyfish to move toward the optimal region. In contrast, a smaller squintability value results in a less distinct separation between desirable and undesirable regions, making it more difficult for other jellyfish to accurately navigate toward the target area. When comparing between multiple optimisers, the smoothness and squintability metrics proposed in this paper are relative and comparison can only be made with common parameters, i.e. the same seed for generating random bases and the same binning width for squintability. Furthermore, the range of the index values should be standardised to the [0, 1] interval as demonstrated for the `holes` and `TIC` indexes.

As a population-based metaheuristic method, JSO requires evaluating multiple jellyfish in each iteration, which increased its computational complexity compared to single-candidate random sampling methods. When choosing optimisers, this higher computational costs needs to be considered. Tuning JSO hyperparameters, such as the number of jellyfish and the maximum number of attempts, can enhance the likelihood of finding the optimal region. However, in some cases, increasing these parameters may not be effective if the index surface is complex or has low squintability. Moreover, the increased computational cost associated with higher parameter can make JSO impractical.

### 2024-08-20 jso.tex
add the discussion section

\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, opacitybacktitle=0.6, colback=white, left=2mm, toprule=.15mm, bottomrule=.15mm, opacityback=0, toptitle=1mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, titlerule=0mm, arc=.35mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, breakable, leftrule=.75mm]

@@

The index properties, including smoothness and squintability, offer
numerical metrics to characterise the complexity of projection pursuit
optimisation problems. Table~\ref{tbl-smoothness-squintability} presents
the parameters for calculating both metrics estimated from the Gaussian
process (variance, range, smooth, and nugget) and the scaled logistic
function (\(\theta_1\) to \(\theta_4\)) for each case considered in the
simulation. The column ``smooth'' is used as the smoothness metric and
the column ``squint'' is calculated as equation
(\ref{eq-squintability-parametric}) as the squintability metric.
Table~\ref{tbl-mod-output} presents the results of fitting a generalised
linear model with a binomial family and a logit link function to a
sample of data in Table~\ref{tbl-mod-data}, where all the three
components (success rate, JS hyper-parameters, and index properties) are
combined. The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the index
properties: smoothness and squintability. Specifically, using 10 more
jellyfish and 10 more tries increases the odd ratio of success by
24.11\% and 11.93\%, respectively. However, being flagged with long
runtime and an increase of data dimension reduce the success rate by
41.72\% and 53.36\%, respectively. The variable \texttt{squintability}
and \texttt{dimension} are significant, suggesting their importance
relative to JSO hyper-parameters in the optimisation success. In
defining a projection pursuit problem, factors such as the
shape-to-find, the index function used, and the data dimension,
determine the properties such as smoothness and squintability. These
metrics can then be compared across different problems to understand
their relative complexites. The results suggest that squintability has a
more significant influence than smoothness on the success rate of JSO.
Once the characteristics of the projection pursuit problem are fully
understood, increasing JSO hyper-parameters can enhance the search
effectiveness, however, it is also important to consider the resulting
increase in computational complexity.

@@

>>>>

\begin{tcolorbox}[enhanced jigsaw, breakable, toptitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, opacityback=0, rightrule=.15mm, bottomrule=.15mm, bottomtitle=1mm, leftrule=.75mm, titlerule=0mm, left=2mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toprule=.15mm, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6]

@@

The index properties, including smoothness and squintability, offer
numerical metrics to characterise the complexity of projection pursuit
optimisation problems. Table~\ref{tbl-smoothness-squintability} presents
the parameters for calculating both metrics estimated from the Gaussian
process (variance, range, smooth, and nugget) and the scaled logistic
function (\(\theta_1\) to \(\theta_4\)) for each case considered in the
simulation. The column ``smooth'' is used as the smoothness metric and
the column ``squint'' is calculated as equation
(\ref{eq-squintability-parametric}) as the squintability metric.
Table~\ref{tbl-mod-output} presents the results of fitting a generalised
linear model with a binomial family and a logit link function to a
sample of data in Table~\ref{tbl-mod-data}, where all the three
components (success rate, JS hyper-parameters, and index properties) are
combined. The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the index
properties: smoothness and squintability. Specifically, using 10 more
jellyfish and 10 more tries increases the odd ratio of success by
24.11\% and 11.93\%, respectively. However, being flagged with long
runtime and an increase of data dimension reduce the success rate by
41.72\% and 53.36\%, respectively. The variable \texttt{squintability}
and \texttt{dimension} are significant, suggesting their importance
relative to JSO hyper-parameters in the optimisation success.

@@

\section{Discussion}\label{sec-discussion}

The two properties, smoothness and squintability, can be calculated once
factors such as the target shape, the chosen index function, and the
data dimension are defined in a projection pursuit problem. In practice,
these metrics can be used to characterise the complexity of the
optimisation task and inform the optimiser selection. For example, a
large smoothness value suggests a smooth index function and both
gradient-based and random sampling optimisers tend to perform well. A
large squintability value indicates a distinct difference in index
values between optimal regions and others. When one jellyfish enters the
optimal region, the high index value it generates will be clearly
distinguishable from spurious values due to this pronounced difference
and lead other jellyfish to move toward the optimal region. In contrast,
a smaller squintability value results in a less distinct separation
between desirable and undesirable regions, making it more difficult for
other jellyfish to accurately navigate toward the target area. When
comparing between multiple optimisers, the smoothness and squintability
metrics proposed in this paper are relative and comparison can only be
made with common parameters, i.e.~the same seed for generating random
bases and the same binning width for squintability. Furthermore, the
range of the index values should be standardised to the {[}0, 1{]}
interval as demonstrated for the \texttt{holes} and \texttt{TIC}
indexes.

As a population-based metaheuristic method, JSO requires evaluating
multiple jellyfish in each iteration, which increased its computational
complexity compared to single-candidate random sampling methods. When
choosing optimisers, this higher computational costs needs to be
considered. Tuning JSO hyperparameters, such as the number of jellyfish
and the maximum number of attempts, can enhance the likelihood of
finding the optimal region. However, in some cases, increasing these
parameters may not be effective if the index surface is complex or has
low squintability. Moreover, the increased computational cost associated
with higher parameter can make JSO impractical.
